### 2016-11-15 greedy_hamiltonian_new_template/chapters/4.sympmor.tex
fist commit

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $(P,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated, by a low dimensional symplectic subspace $(Q,\Omega)$ of dimension $2k$ $(k\ll n)$. We can construct a symplectic basis $A$ for $Q$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have,
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e. $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and therefore does not guarantee conservation of the symplectic structure. On the other hand we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are derived from different Hamiltonians. In the next proposition we show that the loss in the Hamiltonian is constant in time.

@@

\paragraph{\bf Cotangent lift} Suppose that $A$ is of the form
\begin{equation} \label{eq:SyMo:9}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. It is easy to check that $A$ is a symplectic matrix, i.e., $A^T \mathbb J_{2n} A = \mathbb J_{2k}$. The construction of $A$ suggests that the range of $\Phi$ should cover both the potential and the momentum spaces. Hence, we can construct $A$ by forming the combined snapshot matrix
\begin{equation} \label{eq:SyMo:10}
	S_{\text{combined}} = [q_1,\dots,q_n,p_1,\dots,p_n], \qquad \mathbf z_i = (q_i^T,p_i^T)^T,
\end{equation}
and define $\Phi=[u_i,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:9}) cotangent lift minimizes the projection error.

@@

\subsubsection{\bf The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional linear symplectic vector space has a symplectic basis that satisfy conditions (\ref{eq:Hasy:4}). Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \nparallel b$. We seek to grow the size of $S$ by the vectors $a$ and $b$. First we modify $a$ and $b$ to ensure they are symplectic (symplectic orthogonal) to $S$, i.e., we add a linear combination of elements of $S$ to $a$ and $b$, so the resulting vectors are symplectic to $S$. This can be achieved since we assumed $a,b\not \in \text{span}(S)$. So for $a$ we seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation}
	\Omega\left(a + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i f_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation}
	\alpha_i = - \Omega(a,f_i), \quad \beta_i = \Omega(a,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation}
\begin{aligned}
	& \tilde a = a - \sum_{i=1}^k \Omega(a,f_i) e_i + \sum_{i=1}^k \Omega(a,e_i) f_i, \\
	& \tilde b = b - \sum_{i=1}^k \Omega(b,f_i) e_i + \sum_{i=1}^k \Omega(b,e_i) f_i.
\end{aligned}
\end{equation}
Finally we construct the new basis vectors as
\begin{equation}
\begin{aligned}
	& e_{k+1} = \text{sign}(\Omega(\tilde a , \tilde b)) \cdot \frac{\tilde a}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }, \quad f_{k+1} = \frac{\tilde b}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }.
\end{aligned}
\end{equation}
One easily checks that the new basis vectors $e_1,\dots,e_{k+1},f_1,\dots,f_{k+1}$ satisfy all the conditions (\ref{eq:Hasy:4}). Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$. More details about this matrix decomposition are given in the Appendix.

@@

We then propagate (\ref{eq:MoOr:1}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$. The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
\begin{equation}
	z := \underset{s\in S}{\text{argmax }} \| s - AA^+s \|.
\end{equation}
Finally, we update the basis as
\begin{equation}
	e_{k+1} = \tilde z, \quad f_{k+1} = \mathbb J_{2n}^T \tilde z,
\end{equation}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

@@

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Omega_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $f_1 \leftarrow \mathbb J^T_{2n} \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,f_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
\item \hspace{0.5cm}$f_{k+1} \leftarrow \mathbb J^T_{2n} \mathbf z^*$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , f_1,\dots,f_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
\begin{itemize}
\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
\end{itemize}

@@

\begin{definition}
Let $S$ be a subset of $\mathbb R^{2n}$ and $Y_n$ be a general $n$-dimensional subspace of $\mathbb R^{2n}$. The angle between $S$ and $Y_n$ is given by
\begin{equation} \label{eq:new4}
	E(S,Y_n) := \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2.
\end{equation}
The Kolmogorov $n$-width of $S$ in $\mathbb R^n$ is given by
\begin{equation} \label{eq:new5}
	d_{n}(S,\mathbb{R}^n) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
\end{equation}
\end{definition}

@@

We start by symplectifying the vectors provided by the greedy algorithm as
\begin{equation} \label{eq:new6}
\begin{aligned}
	& \xi_1 = e_i, & \bar{\xi}_1 = \mathbb{J}_{2n}^T \xi_1, &\\
	& \xi_i = e_i - P_{2(i-1)} (e_i), & \bar{\xi}_i = \mathbb{J}_{2n}^T, \xi_i &\quad i = 2,3,\dots
\end{aligned}
\end{equation}
Now the projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
\begin{equation} \label{eq:new7}
	P_{2k}(s) = \sum_{i=1}^k \left( \alpha_i(s) \xi_i + \bar{\alpha}_i(s) \bar{\xi}_i \right),
\end{equation}
where $\alpha_i(s)$ and $\bar{\alpha}_i(s)$ for $i=1,\dots,k$ are the expansion coefficients
\begin{equation} \label{eq:new8}
	\alpha_i(s) = - \frac{\Omega(\bar{\xi}_i,s)}{\Omega(\xi_i,\bar{\xi}_i)}, \quad \bar{\alpha}_i(s) = \frac{\Omega(\xi_i,s)}{\Omega(\xi_i,\bar{\xi}_i)},
\end{equation}
for any $s\in S$. Since $\bar{\xi}_i$ is symplectic to span$(A_{2(k-1)})$ we have
\begin{equation} \label{eq:new9}
\begin{aligned}
	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
	&\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
	&= \frac{\| s - P_{2(k-1)}(s) \|_2}{\| e_i - P_{2(k-1)}(e_i) \|_2} \leq 1.
\end{aligned}
\end{equation}
Here, we use the fact that $|\Omega(\xi_i,\bar{\xi}_i)| = |\xi_i^T \mathbb{J}_{2n}\bar{\xi}_i| = \| \xi_i \|^2 = \|\bar{\xi}_i\|^2$ and the last inequality follows from the greedy algorithm which maximizes $e_i$. Similarly we deduce that $|\bar{\alpha}_i(s)|\leq 1$.

@@

Now let $2k$ be the dimension of the desired reduced space. Looking at the definition of Kolmogorov $n$-width we observe that for any $\theta > 1$ we can find a subspace $Y_{2k}$ such that $E(S,Y_{2k}) \leq \theta d_{2k}(S,\mathbb R^n)$. Hence can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that

@@

	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provides exponential approximation properties such that

@@

\subsection{Symplectic Discrete Empirical Interpolation Method (SDEIM)} Consider the Hamiltonian system (\ref{eq:Hasy:7}) and its reduced system (\ref{eq:SyMo:4}) equipped with a symplectic transformation $A$. One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\mathbb R^{n\times n}$ and $\mathbf g$ is a nonlinear function. The reduced system takes the form
\begin{equation} \label{eq:new24}
	\frac{d}{dt} \mathbf y = \underbrace{A^+ \mathbb J_{2n} L A}_{\tilde L} \mathbf y + A^+ \mathbb J_{2n} \mathbf g(A\mathbf y)
\end{equation}
As discussed in Section \ref{chap:MoOr.DEIM:1}, the complexity of evaluating the nonlinear term still depends on $n$, the size of the original system. To overcome this computational bottleneck we use the DEIM approximation for evaluating the nonlinear function $\mathbf g$ as
\begin{equation} \label{eq:new25}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + \underbrace{ A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T \mathbf g(A\mathbf y) }_{\tilde N(\mathbf y)}
\end{equation}
One may argue that if DEIM's projection basis $V$ is rich enough then the approximation of the nonlinear term does not affect the symplectic structure significantly since $\mathbf g$ is already a symplectic function. Nonetheless, for a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system. However, we can guarantee that (\ref{eq:new25}) is a Hamiltonian system by choosing $V=(A^+)^T$. To understand why, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$. Also we have 
\begin{equation} \label{eq:new26}
	\mathbf g(A\mathbf y) = \nabla_{\mathbf z} H_2(\mathbf z) = (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
where the chain rule is used for the second equality. Substituting this into $\tilde N$ we obtain
\begin{equation} \label{eq:new27}
	\tilde N(\mathbf y)= A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T  (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y).
\end{equation}
Taking $V = (A^+)^T$ yields
\begin{equation} \label{eq:new28}
	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T H_2(A \mathbf y) = \mathbb J_{2k} H_2(A \mathbf y),
\end{equation}
using that $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to be a Hamiltonian system.

>>>>

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $(P,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated, by a low dimensional symplectic subspace $(Q,\Omega)$ of dimension $2k$ $(k\ll n)$. We can construct a symplectic basis $A$ for $Q$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e. $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and therefore does not guarantee conservation of the symplectic structure. On the other hand we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are derived from different Hamiltonians. In the next proposition we show that the error in the Hamiltonian is constant in time.

@@

\paragraph{\bf Cotangent lift} Suppose that $A$ is of the form
\begin{equation} \label{eq:SyMo:9}
	A = 
	\begin{pmatrix}
		\Phi & 0 \\
		0 & \Phi
	\end{pmatrix},
\end{equation}
where $\Phi \in \mathbb{R}^{n\times k}$ is an orthonormal matrix. It is easy to check that $A$ is a symplectic matrix, i.e., $A^T \mathbb J_{2n} A = \mathbb J_{2k}$. The construction of $A$ suggests that the range of $\Phi$ should cover both the potential and the momentum spaces. Hence, we can construct $A$ by forming the combined snapshot matrix
\begin{equation} \label{eq:SyMo:10}
	S_{\text{combined}} = [q_1,\dots,q_n,p_1,\dots,p_n], \qquad \mathbf z_i = (q_i^T,p_i^T)^T,
\end{equation}
and define $\Phi=[u_1,\dots,u_k]$, where $u_i$ is the $i$-th left singular vector of $S_{\text{combined}}$. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:9}) cotangent lift minimizes the projection error.

@@

\subsubsection{The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \not \not \parallel b$. We seek to grow the size of $S$ by the vectors $a$ and $b$. First we modify $a$ and $b$ to ensure they are symplectic (symplectic orthogonal) to $S$, i.e., we add a linear combination of elements of $S$ to $a$ and $b$, so the resulting vectors are symplectic to $S$. This can be achieved since we assumed $a,b\not \in \text{span}(S)$. So for $a$ we seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation}
	\Omega\left(a + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i f_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation}
	\alpha_i = - \Omega(a,f_i), \quad \beta_i = \Omega(a,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation}
\begin{aligned}
	& \tilde a = a - \sum_{i=1}^k \Omega(a,f_i) e_i + \sum_{i=1}^k \Omega(a,e_i) f_i, \\
	& \tilde b = b - \sum_{i=1}^k \Omega(b,f_i) e_i + \sum_{i=1}^k \Omega(b,e_i) f_i.
\end{aligned}
\end{equation}
Finally we construct the new basis vectors as
\begin{equation}
\begin{aligned}
	& e_{k+1} = \text{sign}(\Omega(\tilde a , \tilde b)) \cdot \frac{\tilde a}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }, \quad f_{k+1} = \frac{\tilde b}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }.
\end{aligned}
\end{equation}
One easily checks that the new basis vectors $e_1,\dots,e_{k+1},f_1,\dots,f_{k+1}$ satisfy all the conditions (\ref{eq:Hasy:4}). Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$. More details about this matrix decomposition are given in Appendix \ref{chap:Append:1}.

@@

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$. The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
\begin{equation}
	z := \underset{s\in S}{\text{argmax }} \| s - AA^+s \|.
\end{equation}
Finally, we update the basis as
\begin{equation}
	e_{k+1} = \tilde z, \quad f_{k+1} = \mathbb J_{2n}^T \tilde z,
\end{equation}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

@@

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $f_1 \leftarrow \mathbb J^T_{2n} \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,f_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
\item \hspace{0.5cm}$f_{k+1} \leftarrow \mathbb J^T_{2n} \mathbf z^*$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , f_1,\dots,f_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
\begin{itemize}
\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
\end{itemize}

@@

\begin{definition}
Let $S$ be a subset of $\mathbb R^{m}$ and $Y_n$, $n\leq m$, be a general $n$-dimensional subspace of $\mathbb R^{m}$. The angle between $S$ and $Y_n$ is given by
\begin{equation} \label{eq:new4}
	E(S,Y_n) := \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2.
\end{equation}
The Kolmogorov $n$-width of $S$ in $\mathbb R^m$ is given by
\begin{equation} \label{eq:new5}
	d_{n}(S,\mathbb{R}^m) := \inf_{Y_n} E(S,Y_n) = \inf_{Y_n} \sup_{s\in S} \inf_{y\in Y_n} \|s-y\|_2
\end{equation}
\end{definition}

@@

We start by symplectifying the vectors provided by the greedy algorithm as
\begin{equation} \label{eq:new6}
\begin{aligned}
	& \xi_1 = e_i, & \bar{\xi}_1 = \mathbb{J}_{2n}^T \xi_1, &\\
	& \xi_i = e_i - P_{2(i-1)} (e_i), & \bar{\xi}_i = \mathbb{J}_{2n}^T, \xi_i &\quad i = 2,3,\dots
\end{aligned}
\end{equation}
Now the projection of a vector $s\in S$ onto span$(A_{2k})$ can be written using the symplectic basis as
\begin{equation} \label{eq:new7}
	P_{2k}(s) = \sum_{i=1}^k \left( \alpha_i(s) \xi_i + \bar{\alpha}_i(s) \bar{\xi}_i \right),
\end{equation}
where $\alpha_i(s)$ and $\bar{\alpha}_i(s)$ for $i=1,\dots,k$ are the expansion coefficients
\begin{equation} \label{eq:new8}
	\alpha_i(s) = - \frac{\Omega(\bar{\xi}_i,s)}{\Omega(\xi_i,\bar{\xi}_i)}, \quad \bar{\alpha}_i(s) = \frac{\Omega(\xi_i,s)}{\Omega(\xi_i,\bar{\xi}_i)},
\end{equation}
for any $s\in S$. Since $\bar{\xi}_i$ is symplectic to span$(A_{2(k-1)})$ we have
\begin{equation} \label{eq:new9}
\begin{aligned}
	|\alpha_i(s)| = \frac{|\Omega(\bar{\xi}_i,s)|}{|\Omega(\xi_i,\bar{\xi}_i)|} &= \frac{|\Omega( \bar{\xi}_i, s - P_{2(k-1)}(s))|}{|\Omega(\xi_i,\bar{\xi}_i)|} \\
	&\leq \frac{\|\bar{\xi}_i\|_2 \| s - P_{2(k-1)}(s) \|_2}{ \|\xi_i\|_2 \|\bar{\xi}_i\|_2 } \\
	&= \frac{\| s - P_{2(k-1)}(s) \|_2}{\| e_i - P_{2(k-1)}(e_i) \|_2} \leq 1.
\end{aligned}
\end{equation}
Here, we use the fact that $|\Omega(\xi_i,\bar{\xi}_i)| = |\xi_i^T \mathbb{J}_{2n}\bar{\xi}_i| = \| \xi_i \|^2 = \|\bar{\xi}_i\|^2$ and the last inequality follows from the greedy algorithm which maximizes $e_i$. Similarly we deduce that $|\bar{\alpha}_i(s)|\leq 1$.

@@

Now let $2k$ be the dimension of the desired reduced space. Looking at the definition of Kolmogorov $n$-width we observe that for any $\theta > 1$ we can find a subspace $Y_{2k}$ such that $E(S,Y_{2k}) \leq \theta d_{2k}(S,\mathbb R^n)$. Hence we can find vectors $v_1,\dots,v_k,u_1,\dots,u_k\in Y_{2k}$ such that

@@

	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that

@@

\subsection{Symplectic Discrete Empirical Interpolation Method (SDEIM)} Consider the Hamiltonian system (\ref{eq:Hasy:7}) and its reduced system (\ref{eq:SyMo:4}) equipped with a symplectic transformation $A$. One can split the Hamiltonian function $H = H_1 + H_2$ such that $\nabla H_1 = L\mathbf z$ and $\nabla H_2 = \mathbf g(\mathbf z)$, where $L$ is a constant matrix in $\mathbb R^{n\times n}$ and $\mathbf g$ is a nonlinear function. The reduced system takes the form
\begin{equation} \label{eq:new24}
	\frac{d}{dt} \mathbf y = \underbrace{A^+ \mathbb J_{2n} L A}_{\tilde L} \mathbf y + A^+ \mathbb J_{2n} \mathbf g(A\mathbf y)
\end{equation}
As discussed in Section \ref{chap:MoOr.DEIM:1}, the complexity of evaluating the nonlinear term still depends on $n$, the size of the original system. To overcome this computational bottleneck we use the DEIM approximation for evaluating the nonlinear function $\mathbf g$ as
\begin{equation} \label{eq:new25}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + \underbrace{ A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T \mathbf g(A\mathbf y) }_{\tilde N(\mathbf y)}
\end{equation}
One may argue that if DEIM's projection basis $V$ is rich enough then the approximation of the nonlinear term does not affect the symplectic structure significantly since $\mathbf g$ is already a symplectic function. Nonetheless, for a general choice of $V$ the system (\ref{eq:new25}) is not guaranteed to be a Hamiltonian system. However, we can guarantee that (\ref{eq:new25}) is a Hamiltonian system by choosing $V=(A^+)^T$. To understand why, we note that the system (\ref{eq:new25}) is a Hamiltonian system if and only if $\tilde N(\mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} \mathbf g(\mathbf y)$. Also we have 
\begin{equation} \label{eq:new26}
	\mathbf g(A\mathbf y) = \nabla_{\mathbf z} H_2(\mathbf z) = (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
where the chain rule is used for the second equality. Substituting this into $\tilde N$ we obtain
\begin{equation} \label{eq:new27}
	\tilde N(\mathbf y)= A^+ \mathbb J_{2n} V (P^TV)^{-1} P^T  (A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y).
\end{equation}
Taking $V = (A^+)^T$ yields
\begin{equation} \label{eq:new28}
	\tilde N(\mathbf y) = A^+ \mathbb J_{2n}(A^+)^T \nabla_{\mathbf y} H_2(A \mathbf y) = \mathbb J_{2k} \nabla_{\mathbf y} H_2(A \mathbf y),
\end{equation}
using that $(A^+)^T$ is a symplectic matrix. Hence, $V = (A^+)^T$ is a sufficient condition for (\ref{eq:new25}) to be a Hamiltonian system.
